---
title: "Validation of the microclimate model"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# knitr::opts_knit$set(root.dir = "C:/Users/Katerina/Documents/Master thesis/micro")
knitr::opts_knit$set(root.dir = "/Users/usuario/Dropbox/2_supervisions/Katerina/body-temperature/data")
```


### Parameters and running the model

Load packages:

```{r}
library(NicheMapR)
library(microclima)
library(lubridate)
```

Set location (longitude and latitude for mesocosm location, Vila Nova de Gaia) and model parameters. Most model parameters take default values. To improve the fit of the model we decreased the thermal conductivity of the soil (`Thcond` from 2.5 to 1.5) and increased bulk density (`BulkDensity` from 1.3 to 2.45). In addition, we slightly increased the parameter that controls water runon (`rainmult` from 1 to 1.1). We also run the model for two shading levels: full sun (0% shade; `minshade <- 0`) and deep shade (90% shade; `maxshade <- 90`). These shading levels represents the extremes that an organisms may encounter in a given location, as well as all intermediate options (via interpolation).

```{r}
lon <- -8.5898
lat <- 41.1073 
dstart <- "20/05/2022"
dfinish <- "30/06/2022"
minshade <- 0
maxshade <- 30
rainmult <- 1
Thcond <- 2.5
SpecHeat <- 870
Density <- 2.45
BulkDensity <- 1.3
windfac <- 1
REFL <- 0.2
cap <- FALSE
SLE <- 0.95
warm <- 0
Usrhyt <- 0.01
clearsky <- FALSE
soilgrids <- 0
spatial <- NA
ERR <- 1
```

We obtained a digital elevation model (DEM) for the study location using the `get_dem` function of the microclima R-package (citation), and downloaded it for later use. From the DEM we calculated elevation, slope and aspect, as well as the horizon angles, which impact on the incidence of solar radiation during the day.

```{r}
# dem <- microclima::get_dem(r = NA, lat = lat, lon = lon, resolution = 30, 
#                             zmin = -20, xdims = 100, ydims = 100)
# save(dem, file = "dem.Rda")

load("dem.Rda")

elev <- raster::extract(dem, cbind(lon, lat))[1]        
xy <- data.frame(x = lon, y = lat)
sp::coordinates(xy) = ~x + y
sp::proj4string(xy) = "+init=epsg:4326"
xy <- as.data.frame(sp::spTransform(xy, raster::crs(dem)))
slope <- raster::terrain(dem, unit = "degrees")
slope <- raster::extract(slope, xy)
aspect <- raster::terrain(dem, opt = "aspect", unit = "degrees")
aspect <- raster::extract(aspect, xy)
ha36 <- 0
for (i in 0:35) {
  har <- microclima::horizonangle(dem, i * 10, raster::res(dem)[1])
  ha36[i + 1] <- atan(raster::extract(har, xy)) * (180/pi)
}
hori <- spline(x = ha36, n = 24, method =  'periodic')$y
hori[hori < 0] <- 0
hori[hori > 90] <- 90
```

With these parameters we run the model with the gads (`run.gads = 1`), soil moisture (`runmoist = 1`), and snow models (`snowmodel = 1`) turned on. In a first run, we set `save` to 1 to download environmental forcing variables from NCEP and store them locally. Then, we set `save = 2` to read in saved weather data in next runs (e.g. while testing parameter values). 

```{r}
micro <- micro_ncep(loc = c(lon, lat), SLE = SLE, warm = warm, soilgrids = soilgrids, 
               dstart = dstart, dfinish = dfinish, Usrhyt = Usrhyt,
               slope = slope, aspect = aspect, REFL = REFL,
               hori = hori, minshade = minshade, maxshade = maxshade, 
               rainmult = rainmult, BulkDensity =  BulkDensity, cap = cap,
               Density = Density, Thcond = Thcond, SpecHeat = SpecHeat,
               windfac = windfac, spatial = spatial, ERR = ERR, dem = dem, 
               runshade = 1, run.gads = 1, snowmodel = 1, runmoist = 1,
               save = 2)
```

```{r, echo=F, eval=F}
# save(micro, file = "micro_ncep.Rda")
load("micro_ncep.Rda")
```

Retrieve output for soil temperatures at full sun (`soil`), soil temperatures at maximum shade (`shadsoil`), relative humidity in soil at full sun (`humid`) and relative humidity in soil at maximum shade (`shadhumid`). Also output for micrometeorological conditions under minimum shade (`metout`) and maximum shade (`shadmet`).

```{r}
metout <- data.frame(micro$metout)
shadmet <- data.frame(micro$shadmet)
soil <- data.frame(micro$soil) # soil temperatures at full sun
shadsoil <- data.frame(micro$shadsoil) # soil temperatures at maximum shade
humid <- data.frame(micro$humid) # RH in soil at full sun
shadhumid <- data.frame(micro$shadhumid) # RH in soil at maximum shade
```

### Temperature at full sun

```{r, echo=F}

# location temperature (rock)
tloc_sun_0cm <- read.csv(file = 'MS004T.csv', head=T)
tloc_sun_0cm <- data.frame(datetime = round(as.POSIXct(strptime(tloc_sun_0cm[,1], "%d/%m/%Y %H:%M")),"hours"),temp = as.numeric(paste(tloc_sun_0cm[,3],tloc_sun_0cm[,4], sep='.')))
year(tloc_sun_0cm$datetime) <- 2022

# soil sun
tsoil_sun_0cm <- read.csv(file = 'MS002T.csv', head=T)
tsoil_sun_0cm <- data.frame(datetime = round(as.POSIXct(strptime(tsoil_sun_0cm[,1], "%d/%m/%Y %H:%M")),"hours"),temp = as.numeric(paste(tsoil_sun_0cm[,3],tsoil_sun_0cm[,4], sep='.')))
year(tsoil_sun_0cm$datetime) <- 2022

# soil sun buried (15 cm)
tsoil_sun_buried <- read.csv(file = 'MS005T.csv', head=T)
tsoil_sun_buried <- data.frame(datetime = round(as.POSIXct(strptime(tsoil_sun_buried[,1], "%d/%m/%Y %H:%M")),"hours"),temp = as.numeric(paste(tsoil_sun_buried[,3],tsoil_sun_buried[,4], sep='.')))
year(tsoil_sun_buried$datetime) <- 2022


```

```{r, out.width="100%", echo=F}
source("../code/aux_functions.R") # To calculate RMSEs

# plot
par(mfrow=c(3,1), mar=c(2.5, 4, 0.5, 0.1))

obs_range <- as.POSIXct(c("2022-05-20 00:00:00 UTC","2022-06-30 23:00:00 UTC"))

with(metout, plot(micro$dates, TALOC, type='l', 
                xlim=obs_range, ylim=c(5,55),
                xlab='', ylab='Trock sun 0 cm (ºC)'))
with(tloc_sun_0cm, points(datetime, temp, type='l', col='red'))
text(as.POSIXct("2022-06-01 00:00:00 UTC"), max(metout$TALOC) * 0.9,
     paste0("RMSE = ", rmse.calc(metout, tloc_sun_0cm, depth=1, micro)))



with(soil, plot(micro$dates, D0cm, type='l', 
                xlim=obs_range, ylim=c(5,55),
                xlab='', ylab='Tsoil sun 0 cm (ºC)'))
with(tsoil_sun_0cm, points(datetime, temp, type='l', col='red'))
text(as.POSIXct("2022-06-01 00:00:00 UTC"), max(soil$D0cm) * 0.8,
     paste0("RMSE = ", rmse.calc(soil, tsoil_sun_0cm, depth=1, micro)))


with(soil, plot(micro$dates, D15cm, type='l', 
                xlim=obs_range, ylim=c(5,35),
                xlab='', ylab='Tsoil sun 15 cm (ºC)'))
with(tsoil_sun_buried, points(datetime, temp, type='l', col='red'))
text(as.POSIXct("2022-06-02 00:00:00 UTC"), max(soil$D15cm) * 0.95,
     paste0("RMSE = ", rmse.calc(soil, tsoil_sun_buried, depth=7, micro)))



```

### Temperature at shade (90%)

```{r, out.width="100%", echo=F}
# soil temperature
# air temperature
tair_shad_0cm <- read.csv(file = 'MS006T.csv', head=T)
tair_shad_0cm <- data.frame(datetime = round(as.POSIXct(strptime(tair_shad_0cm[,1], "%d/%m/%Y %H:%M")),"hours"),temp = as.numeric(paste(tair_shad_0cm[,3],tair_shad_0cm[,4], sep='.')))
year(tair_shad_0cm$datetime) <- 2022

# rock shadow
tloc_shad_0cm <- read.csv(file = 'MS003T.csv', head=T)
tloc_shad_0cm <- data.frame(datetime = round(as.POSIXct(strptime(tloc_shad_0cm[,1], "%d/%m/%Y %H:%M")),"hours"),temp = as.numeric(paste(tloc_shad_0cm[,3],tloc_shad_0cm[,4], sep='.')))
year(tloc_shad_0cm$datetime) <- 2022

# soil shadow
tsoil_shad_0cm <- read.csv(file = 'MS001T.csv', head=T)
tsoil_shad_0cm <- data.frame(datetime = round(as.POSIXct(strptime(tsoil_shad_0cm[,1], "%d/%m/%Y %H:%M")),"hours"),temp = as.numeric(paste(tsoil_shad_0cm[,3],tsoil_shad_0cm[,4], sep='.')))
year(tsoil_shad_0cm$datetime) <- 2022


# plot
par(mfrow=c(3,1), mar=c(2.5, 4, 0.5, 0.1))

obs_range <- as.POSIXct(c("2022-05-20 00:00:00 UTC","2022-06-30 23:00:00 UTC"))

with(shadmet, plot(micro$dates, TALOC, type='l', 
                xlim=obs_range, ylim=c(10,35),
                xlab='', ylab='Tair shadow 0 cm (ºC)'))
with(tair_shad_0cm, points(datetime, temp, type='l', col='red'))
text(as.POSIXct("2022-05-30 00:00:00 UTC"), max(shadmet$TAREF) * 0.65,
     paste0("RMSE = ", rmse.calc(shadmet, tair_shad_0cm, depth=2, micro)))

with(shadmet, plot(micro$dates, TALOC, type='l', 
                xlim=obs_range, ylim=c(10,35),
                xlab='', ylab='Trock shadow 0 cm (ºC)'))
with(tloc_shad_0cm, points(datetime, temp, type='l', col='red'))
text(as.POSIXct("2022-05-30 00:00:00 UTC"), max(shadmet$TALOC) * 0.6,
     paste0("RMSE = ", rmse.calc(shadmet, tloc_shad_0cm, depth=1, micro)))


with(shadsoil, plot(micro$dates, D0cm, type='l', 
                xlim=obs_range, ylim=c(10,35),
                xlab='', ylab='Tsoil shadow 0 cm (ºC)'))
with(tsoil_shad_0cm, points(datetime, temp, type='l', col='red'))
text(as.POSIXct("2022-05-30 00:00:00 UTC"), max(shadsoil$D0cm) * 0.6,
     paste0("RMSE = ", rmse.calc(shadsoil, tsoil_shad_0cm, depth=1, micro)))


```

### Humidity at full sun

```{r, out.width="100%", echo=F}

# location humidity (rock)
hloc_sun_0cm <- read.csv(file = 'MS004H.csv', head=T)
hloc_sun_0cm <- data.frame(datetime = round(as.POSIXct(strptime(hloc_sun_0cm[,1], "%d/%m/%Y %H:%M")),"hours"),hum = as.numeric(paste(hloc_sun_0cm[,3],hloc_sun_0cm[,4], sep='.')))
year(hloc_sun_0cm$datetime) <- 2022
hloc_sun_0cm$hum[hloc_sun_0cm$hum > 100] <- 100

# humidity soil sun
hsoil_sun_0cm <- read.csv(file = 'MS002H.csv', head=T)
hsoil_sun_0cm <- data.frame(datetime = round(as.POSIXct(strptime(hsoil_sun_0cm[,1], "%d/%m/%Y %H:%M")),"hours"),hum = as.numeric(paste(hsoil_sun_0cm[,3],hsoil_sun_0cm[,4], sep='.')))
year(hsoil_sun_0cm$datetime) <- 2022
hsoil_sun_0cm$hum[hsoil_sun_0cm$hum > 100] <- 100

# humidity soil sun buried (? cm)
hsoil_sun_buried <- read.csv(file = 'MS005H.csv', head=T)
hsoil_sun_buried <- data.frame(datetime = round(as.POSIXct(strptime(hsoil_sun_buried[,1], "%d/%m/%Y %H:%M")),"hours"),hum = as.numeric(paste(hsoil_sun_buried[,3],hsoil_sun_buried[,4], sep='.')))
year(hsoil_sun_buried$datetime) <- 2022
hsoil_sun_buried$hum[hsoil_sun_buried$hum > 100] <- 100


# plot

# plot
par(mfrow=c(3,1), mar=c(2.5, 4, 0.5, 0.1))

obs_range <- as.POSIXct(c("2022-05-20 00:00:00 UTC","2022-06-30 23:00:00 UTC"))

with(metout, plot(micro$dates, RHLOC, type='l', 
                xlim=obs_range, ylim=c(0,100),
                xlab='', ylab='RH rock sun 0 cm (ºC)'))
with(hloc_sun_0cm, points(datetime, hum, type='l', col='red'))
text(as.POSIXct("2022-06-01 00:00:00 UTC"), max(metout$RHLOC) * 0.2,
     paste0("RMSE = ", rmse.calc(metout, hloc_sun_0cm, depth=3, micro)))



with(humid, plot(micro$dates, RH0cm * 100, type='l', 
                xlim=obs_range, ylim=c(0,100),
                xlab='', ylab='RH soil sun 0 cm (ºC)'))
with(hsoil_sun_0cm, points(datetime, hum, type='l', col='red'))
text(as.POSIXct("2022-05-31 00:00:00 UTC"), max(humid$RH0cm) * 0.2,
     paste0("RMSE = ", rmse.calc(humid, hsoil_sun_0cm, depth=1, micro)))


with(humid, plot(micro$dates, RH15cm * 100, type='l', 
                xlim=obs_range, ylim = c(0, 100),
                xlab='', ylab='RH soil sun 30 cm (ºC)'))
with(hsoil_sun_buried, points(datetime, hum, type='l', col='red'))
text(as.POSIXct("2022-06-07 00:00:00 UTC"), max(humid$RH15cm) * 0.8,
     paste0("RMSE = ", rmse.calc(humid, hsoil_sun_buried, depth=7, micro)))
```

### Humidity at shade (90%)

```{r, out.width="100%", echo=F}

# air humidity
hair_shad_0cm <- read.csv(file = 'MS006H.csv', head=T)
hair_shad_0cm <- data.frame(datetime = round(as.POSIXct(strptime(hair_shad_0cm[,1], "%d/%m/%Y %H:%M")),"hours"),hum = as.numeric(paste(hair_shad_0cm[,3],hair_shad_0cm[,4], sep='.')))
year(hair_shad_0cm$datetime) <- 2022

# rock shadow
hloc_shad_0cm <- read.csv(file = 'MS003H.csv', head=T)
hloc_shad_0cm <- data.frame(datetime = round(as.POSIXct(strptime(hloc_shad_0cm[,1], "%d/%m/%Y %H:%M")),"hours"),hum = as.numeric(paste(hloc_shad_0cm[,3],hloc_shad_0cm[,4], sep='.')))
year(hloc_shad_0cm$datetime) <- 2022

# soil shadow
hsoil_shad_0cm <- read.csv(file = 'MS001H.csv', head=T)
hsoil_shad_0cm <- data.frame(datetime = round(as.POSIXct(strptime(hsoil_shad_0cm[,1], "%d/%m/%Y %H:%M")),"hours"),hum = as.numeric(paste(hsoil_shad_0cm[,3],hsoil_shad_0cm[,4], sep='.')))
year(hsoil_shad_0cm$datetime) <- 2022
hsoil_shad_0cm$hum <- hsoil_shad_0cm$hum/100


# plot
par(mfrow=c(3,1), mar=c(2.5, 4, 0.5, 0.1))

obs_range <- as.POSIXct(c("2022-05-20 00:00:00 UTC","2022-06-30 23:00:00 UTC"))

with(shadmet, plot(micro$dates, RH, type='l', 
                xlim=obs_range, ylim=c(0,100),
                xlab='', ylab='H air shadow 0 cm (ºC)'))
with(hair_shad_0cm, points(datetime, hum, type='l', col='red'))
text(as.POSIXct("2022-06-04 00:00:00 UTC"), max(shadmet$RH) * 0.6,
     paste0("RMSE = ", rmse.calc(shadmet, hair_shad_0cm, depth=4, micro)))

with(shadmet, plot(micro$dates, RHLOC, type='l', 
                xlim=obs_range, ylim=c(0,100),
                xlab='', ylab= 'H rock shadow 0 cm (ºC)'))
with(hloc_shad_0cm, points(datetime, hum, type='l', col='red'))
text(as.POSIXct("2022-06-04 00:00:00 UTC"), max(shadmet$RHLOC) * 0.5,
     paste0("RMSE = ", rmse.calc(shadmet, hloc_shad_0cm, depth=3, micro)))


with(shadhumid, plot(micro$dates, RH0cm, type='l', 
                xlim=obs_range, ylim=c(0,1),
                xlab='', ylab='H soil shadow 0 cm (ºC)'))
with(hsoil_shad_0cm, points(datetime, hum, type='l', col='red'))
text(as.POSIXct("2022-06-01 00:00:00 UTC"), max(shadhumid$RH0cm) * 0.7,
     paste0("RMSE = ", rmse.calc(shadhumid, hsoil_shad_0cm, depth=1, micro)))
```






